Staged Abstract Interpreters

### 2019-08-16 paper/10_intro.tex
update and spaces

Recent progress on methodologies such as Abstracting Abstract
Machine (AAM) \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10}
uncovers an approach to constructing abstract interpreters from
systematic principles, rather than ad-hoc engineering.  By deriving
semantics artifacts for abstract interpretation from their concrete
counterparts (for example, abstract machines), soundness can be
easily established by examining the transformation of semantic
artifacts.
This systematic abstraction approach can be tailored to different
language features (such as state, first-class control, exception, and
concurrency) and sensitivity analyses \cite{DBLP:conf/icfp/Gilray0M16,
  DBLP:conf/popl/GilrayL0MH16, Darais:2015:GTM:2814270.2814308}. It
has also been applied to various small-step abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
  Sergey:2013:MAI:2491956.2491979} and big-step definitional
interpreters \cite{Wei:2018:RAA:3243631.3236800,
  DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

Based on the idea of abstracting abstract machines, more
pragmatically, several implementation strategies utilizing purely
functional programming techniques to build abstract interpreters have
emerged. Such techniques include monads, monad transformers
\cite{DBLP:journals/pacmpl/DaraisLNH17,
  Sergey:2013:MAI:2491956.2491979}, arrows
\cite{Keidel:2018:CSP:3243631.3236767}, extensible effects
\cite{Kiselyov:2015:FMM:2804302.2804319}, etc. The pure approaches provide certain
benefits, including that the abstract interpretation artifacts can be
built compositionally and modularly, e.g., by using monad
transformers. Therefore, the soundness of analysis can be proven with
less effort, whether by mechanized proofs
\cite{Darais:2016:CGC:2951913.2951934} or paper-based proofs
\cite{Keidel:2018:CSP:3243631.3236767}.  Also, referential
transparency and equational reasoning allow programmers to
confidently reason about the correctness of their implementations.

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes the program by traversing the abstract syntax tree (AST), which incurs
interpretive overhead, such as the pattern matching on the ASTs and recursive
calls on the subexpressions. Such overhead can be negligible if the abstract
interpreter only runs on the program a single time, but can accumulate
significantly if run repeatedly (for example, on libraries).  Second, it is
common that abstract interpreters written in pure languages extensively use
effect systems to implement the semantics of abstract interpretation. For
example, the collecting interpreter that returns all possible runtime values is
nondeterministic.  Although such pure approaches have their own merits, compared
with imperative, stateful, implementations, they are significantly slower. In
order to reconcile these approaches, one must use a programming technique which
allows for high-level, functional abstractions during implementation, but which
bypasses the incurred overhead at runtime.

Roughly at the same time when abstract interpretation was proposed in
1970s, \citet{futamura1971partial, Futamura1999} revealed a close
connection between interpreters and compilers through a hierarchy of
specializations, namely, Futamura Projections. The first Futamura
projection shows that specializing an interpreter to the input program
removes interpretation overhead and yields an equivalent executable
that runs faster.  The challenge here: can we apply the same idea to
general abstract interpreters to make hem run faster?
To achieve this, we need to handle a nonstandard semantics that
extensively uses nondeterminism, operates on abstract domains, and
computes fixed-point iteratively with a cache.

In this paper, we show that the first Futamura projection can be naturally
extended to abstract interpreters. We present an abstraction-without-regret
framework to eliminate performance penalties for abstract interpreters, while
keeping the benefits inherent in purely functional programming. In short, our
approach borrows ideas and techniques from multi-stage programming and embedded
DSLs to abstract interpreters:  1) To remove the overhead from interpretation
and effect layers, we specialize the abstract interpreter with respect to an
input program and then generate efficient low-level code that does the actual
analysis.  2) Inspired by the tagless-final interpreters
\cite{DBLP:journals/jfp/CaretteKS09}, we construct a generic interpreter that
abstracts over binding-time and different semantics, which allows user to
implement different semantics modularly, including the staged abstract
interpretation semantics. As a result, we can derive the staged abstract
interpreters without intrusive changes to its corresponding unstaged
counterpart, and also preserve the soundness of analysis. In this sense, our
approach allows for no regret for both performance and engineering effort. We
elaborate these two main ideas in detail, as follows.

@@

The first Futamura projection specifically shows that specializing an
interpreter w.r.t. the input program yields an equivalent
executable. For instance,
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{eval}
is an interpreter for some language:
\begin{lstlisting}
  def eval(e: Expr)(arg: Input): Value
\end{lstlisting}
Given a program $e_0$ of type
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Expr},
by applying the specialization, we can obtain a specialized
interpreter $\texttt{eval}_{\texttt{e0}}$ :
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Input}
$\to$
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Value},
which is the so-called \textit{equivalent executable}.

@@

Partial evaluation \cite{DBLP:books/daglib/0072559} was the first proposed
approach to realize Futamura projections: it first discovers the binding-times
(\textit{static} or \textit{dynamic}) of variables, then evaluates the static
part of the program, and generates a residualized program that solely relies on
the dynamic part. However, it is hard to precisely analyze binding-times given
an arbitrary program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} lets users explicitly program over
a two-level semantics \cite{NIELSON198859}, by manually annotating the
binding-times of variables in the program. Then, the MSP system will check
whether these annotations are consistent, and specialize the program using that
information. The staging annotations can be either syntactic (e.g., quote and
quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight Modular
Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

Similar to two-level semantics \cite{NIELSON1989117}, we found
monads provide a good abstraction to hide the detail of abstract
interpretation semantics, and allows us to fit code generation into
the semantics. We use \citet{DBLP:journals/pacmpl/DaraisLNH17}'s
abstract definitional interpreter that uses monad transformers as the
unstaged baseline.  Our proposed framework adopts the type-based
multi-stage programming from the Lightweight Modular Staging framework,
and implements the Futamura projection of a big-step abstract
interpreter for a small, stateful, higher-order language. By deriving
staged monads that can be used to do code generation, we obtain a
staged abstract interpreter that fulfills the specialization. After
staging, the generated code is specialized to the input program, and
all monadic operations are inlined and compiled down to low-level
Scala code.

@@

Although the four artifacts may look like dissimilar at first
glance, they in fact are all firmly rooted in the concrete semantics of
the language.  This observation provides a way to abstract over the
interpreter and achieve the flexibility of reinterpreting the shared
interpreter. Generic interpreters using monads have already been
applied to abstract interpreters; we further introduce the
binding-time abstraction into the generic interpreter.  The generic
interpreter returns a value of monadic type, which can be varied by
different semantics. The domain of the interpreter and the effects
such as read, state, and nondeterminism can all be wrapped into this
monadic type.  The binding-time abstraction is represented by a
higher-kinded type, and controls whether the interpreter produces
values directly or generates code. It is worth mentioning that the
binding-time type is also injected into the monadic type, so that we
will distinguish normal monads and staged monads.

>>>>

Recent progress on methodologies such as Abstracting Abstract Machine (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} uncovers an
operational approach to constructing abstract interpreters.  By deriving
semantics artifacts for abstract interpretation from their concrete
counterparts (for example, abstract machines), soundness can be easily
established by examining the transformation of semantic artifacts.
This systematic abstraction approach can be tailored to different
language features (such as state, first-class control, exception, and
concurrency) and sensitivity analyses \cite{DBLP:conf/icfp/Gilray0M16,
  DBLP:conf/popl/GilrayL0MH16, Darais:2015:GTM:2814270.2814308}. It
has also been applied to various small-step abstract machines
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
  Sergey:2013:MAI:2491956.2491979} and big-step definitional
interpreters \cite{Wei:2018:RAA:3243631.3236800,
  DBLP:journals/pacmpl/DaraisLNH17, Keidel:2018:CSP:3243631.3236767}.

Based on the idea of abstracting abstract machines, more
pragmatically, several implementation strategies utilizing purely
functional programming techniques to build abstract interpreters have
emerged. Such techniques include monads, monad transformers
\cite{DBLP:journals/pacmpl/DaraisLNH17,
  Sergey:2013:MAI:2491956.2491979}, arrows
\cite{Keidel:2018:CSP:3243631.3236767}, extensible effects
\cite{Kiselyov:2015:FMM:2804302.2804319, Githubsemantic}, etc. The pure
approaches provide certain benefits, including that the abstract interpretation
artifacts can be built compositionally and modularly, e.g., by using monad
transformers. Therefore, the soundness of analysis can be proven with less
effort, whether by mechanized proofs \cite{Darais:2016:CGC:2951913.2951934} or
paper-based proofs \cite{Keidel:2018:CSP:3243631.3236767}.  Also, referential
transparency and equational reasoning allow programmers to confidently reason
about the correctness of their implementations.

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes the program by traversing the abstract syntax tree (AST), which incurs
interpretive overhead, such as the pattern matching on the ASTs and recursive
calls on the subexpressions. Such overhead can be negligible if the abstract
interpreter only runs on the program a single time, but can accumulate
significantly if run repeatedly (for example, on libraries).  Second, it is
common that abstract interpreters written in pure languages extensively use
functional encodings of side-effects to implement the semantics of abstract interpretation. 
For example, the collecting interpreter that returns all possible runtime values is
nondeterministic, which can be modeled by a list monad.  Although such pure
approaches have their own merits, compared with imperative, stateful,
implementations, they are significantly slower.
In order to reconcile these approaches, one must use a programming technique
which allows for high-level abstractions during implementation, but which
bypasses the incurred overhead at runtime.

Roughly at the same time when abstract interpretation was proposed in
1970s, \citet{futamura1971partial, Futamura1999} revealed a close
connection between interpreters and compilers through a hierarchy of
specializations, namely, Futamura Projections. The first Futamura
projection shows that specializing an interpreter to the input program
removes interpretation overhead and yields an equivalent executable
that runs faster. The idea of Futamura projection has been applied to 
other similar performance problems and shown success, examples including
query compilers\cite{DBLP:conf/sigmod/TahboubER18}, PyPy \footnote{PyPy:
\url{http://pypy.org/}}, Truffle \cite{Marr:2015:TVP:2814270.2814275} and etc.
The challenge here: can we apply the same idea to general abstract interpreters
to make hem run faster? To achieve this, we need to specialize a nonstandard
semantics that extensively uses nondeterminism, operates on abstract domains,
and computes fixed-point iteratively with a cache. The solution is not simply obivious.

In this paper, we show that the first Futamura projection can be naturally
extended to abstract interpreters. We present an abstraction-without-regret
framework to eliminate performance penalties for abstract interpreters, while
keeping the implementation as close to its conceptual model.
In short, our approach borrows ideas and techniques from multi-stage
programming and embedded DSLs to abstract interpreters:  1) To remove the
overhead from interpretation and effect layers, we specialize the abstract
interpreter via staging and then generate efficient low-level code that does
the actual analysis. 2) Inspired by the tagless-final interpreters
\cite{DBLP:journals/jfp/CaretteKS09}, we construct a generic interpreter that
abstracts over binding-time and different semantics, which allows the staged
abstract interpreter can be derived from its unstaged counterpart.
As a result, the derived staged abstract interpreters has no intrusive changes
to the underlying abstract semantics, therefore preserves the soundness.
In this sense, our approach allows for no regret for both performance and
engineering effort. We elaborate these two main ideas in detail, as follows.

@@

The first Futamura projection specifically shows that specializing an
interpreter w.r.t. the input program yields an equivalent
executable. For instance,
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{eval : Expr -> Input -> Value}
is an interpreter for some language. Given a program $e_0$ of type
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Expr},
by applying the specialization, we can obtain a specialized
interpreter $\texttt{eval}_{\texttt{e0}}$ :
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{Input -> Value},
which is the so-called \textit{equivalent executable}.

@@

The first proposed approach to realize Futamura projections is partial
evaluation \cite{DBLP:books/daglib/0072559}, which relies on an automatic
binding-time analysis (\textit{static} or \textit{dynamic}).  The underlying
representation can be viewed as \textit{two-level semantics}
\cite{NIELSON1989117, NIELSON198859} that distinguishes compile-time
computation and run-time computation.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} lets users explicitly program over
a two-level syntax \cite{Nielson:1992:TFL:130665}, by manually
annotating the binding-times of variables in the program. Then, the MSP system
will check whether these annotations are consistent, and specialize the program
using that information. The staging annotations can be either syntactic (e.g.,
quote and quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight
Modular Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

Similar to two-level semantics \cite{NIELSON1989117}, we found
monads provide a good abstraction to hide the detail of abstract
interpretation semantics, and allows us to fit code generation into
the semantics. We use \citet{DBLP:journals/pacmpl/DaraisLNH17}'s
abstract definitional interpreter that uses monad transformers as the
unstaged baseline. However, the monads do not have a clear stage distinction.
We further introduce binding-time and make the monads binding-time polymorphic.
Our proposed framework adopts the type-based multi-stage programming from the
Lightweight Modular Staging framework, and implements the Futamura projection
of a big-step abstract interpreter for a small, stateful, higher-order
language. By deriving staged monads that can be used to do code generation, we
obtain a staged abstract interpreter that can be used for specialization. After
staging, the generated code is specialized to the input program, and all
monadic operations are inlined and compiled down to low-level Scala code.

@@

Although the four artifacts may look like dissimilar at first glance, they in
fact are all firmly rooted in the concrete semantics of the language.  This
observation provides a way to abstract over the interpreter and achieve the
flexibility of reinterpreting the shared interpreter.  The generic interpreter
returns a value of monadic type, which can be varied by different semantics.
The domain of the interpreter and the effects such as read, state, and
nondeterminism can all be wrapped into this monadic type.  The binding-time
abstraction is represented by a higher-kinded type, and controls whether the
interpreter produces values directly or generates code. It is worth mentioning
that the binding-time type is also injected into the monadic type, so that we
will distinguish normal monads and staged monads.

### 2019-08-16 paper/50_stageabs.tex
update and spaces

The \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{ap_clo} method is also similar: we use the staged version of \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{lift_nd} to
lift the next-stage set of closures into the monad stack, and generate a
next-stage value for each closure \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{clo} in the set, by using \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{emit_ap_clo}. The
method \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{emit_ap_clo} generates the current-stage representation of the
future-stage result of application.
Finally, we put the future \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{out} cache, store, and values back into
the current-stage monads.
\begin{lstlisting}
  def emit_ap_clo(rator: Rep[AbsValue], rand: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(Set[ValSt], Cache)]
  def ap_clo(ev: Expr => Ans)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    σ   <- get_store;      clo <- lift_nd[AbsValue](rator)
    in  <- ask_in_cache;   out <- get_out_cache
    res <- lift_nd[(Set[ValSt], Cache)](Set(emit_ap_clo(clo, arg, σ, in, out)))
    _   <- put_out_cache(res._2)
    vs  <- lift_nd[ValSt](res._1)
    _   <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

% Monad is the bridge that connects future-stage values

@@

%\paragraph{Heterogeneous Staging} The generated program is in A-Normal form. \todo{TODO}

@@

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters. In this section, we review and summarize our recipe
to achieve the staged abstract interpreter, discuss correctness issues,
and different design choices.

@@

%\todo{maybe show a small example?}

>>>>

The \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{ap_clo} method is also similar: we use the staged version of \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{lift_nd} to
lift the next-stage set of closures into the monad stack, and generate a
next-stage value for each closure \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{clo} in the set, by using \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{emit_ap_clo}. The
method \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{emit_ap_clo} generates the current-stage representation of the
future-stage result of application.
Finally, we put the future \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{out} cache, store, and values back into
the current-stage monads.
\begin{lstlisting}
  def emit_ap_clo(rator: Rep[AbsValue], rand: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(Set[ValSt], Cache)]
  def ap_clo(ev: Expr => Ans)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    σ   <- get_store;      clo <- lift_nd[AbsValue](rator)
    in  <- ask_in_cache;   out <- get_out_cache
    res <- lift_nd[(Set[ValSt], Cache)](Set(emit_ap_clo(clo, arg, σ, in, out)))
    _   <- put_out_cache(res._2);  vs <- lift_nd[ValSt](res._1);  _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

@@

@@

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters from an \textit{operational} perspective.
In this section, we review and summarize our recipe to achieve the staged
abstract interpreter, discuss correctness issues, and different design choices.

@@

### 2019-08-16 paper/60_cases.tex
update and spaces

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Control-flow Analysis} \label{cfa}

@@

\begin{figure}[h!]
\begin{lstlisting}
  def fix_gc(ev: (Expr => Ans) => (Expr => Ans))(e: Expr): Ans = for {
    ψ <- ask_roots ... // omit code for retrieving ρ, σ, in and out
    val res: Rep[(Set[(Value,Store)], Cache)] = ... // omit code for checking caches
    _ <- put_out_cache(res._2)
    vs <- lift_nd(res._1)
    σ <- gc(ψ ⊔ root(vs._2)) // perform abstract garbage collection
    _ <- put_store(σ)
  } yield v
\end{lstlisting}
\vspace{-2em}
\caption{\texttt{fix\_gc} for abstract garbage collection} \label{fixgc}
\end{figure}

@@

\subsection{Modular Staged Analysis for Free} \label{modular}

% using different k?

@@

Consider there is a library that contains a list map function \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{map}, 
\begin{lstlisting}
  (define (map xs f)
    (if (null? xs) '() (cons (f (car xs)) (map (cdr xs) f))))
\end{lstlisting}

The specialization (0CFA and without store-widening) of \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{close} on
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{map} generates the following next-stage Scala code, that is, the
function \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{x39}:
\begin{lstlisting}
  val x39 = {(x40: List[Set[AbsValue]],          // argument value list
              x41: Map[Addr, Set[AbsValue]],     // store
              x42: Map[Config, Set[ValSt]],      // in cache
              x43: Map[Config, Set[ValSt]]) =>   // out cache
    val x50 = List[Addr](ZCFAAddr("xs"), ZCFAAddr("f")) // addresses of arguments
    val x51 = x50.zip(x40)                              // addresses paired with their values
    val x61 = x51.foldLeft (x41) { case (x52, x53) =>   // prepare a new store that joins
      val x54 = x53._1;  val x55 = x53._2               //   the arguments and their values
        val x58 = x52.getOrElse(x54, x57)               //   into the latest store, initially x41
        val x59 = x58.union(x55)
        x52 + (x54 -> x59)
    } ...
  }
\end{lstlisting}

We borrow the notation traditional modular analysis and realize that
such generated function is a reusable artifact which can be used
modularly.  The generated functions already eliminate the
interpretation overhead and monadic overhead, therefore can be
consider as partial summaries. Although, we still need to run the
generated function at next stage to obtain the analysis result.  At
next stage, we can provide different abstract arguments, store and
caches, depending on the context where it is called, thus the analysis
is modular.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{-2em}
\subsection{Control-flow Analysis} \label{cfa}

@@

\vspace{-1em}
\begin{figure}[h!]
\begin{lstlisting}
  def fix_gc(ev: (Expr => Ans) => (Expr => Ans))(e: Expr): Ans = for {
    ψ <- ask_roots ... // omit code for retrieving ρ, σ, in and out
    val res: Rep[(Set[(Value,Store)], Cache)] = ...  // omit code for checking caches
    _ <- put_out_cache(res._2);  vs <- lift_nd(res._1)
    σ <- gc(ψ ⊔ root(vs._2));     _ <- put_store(σ)  // gc performs abstract garbage collection
  } yield v
\end{lstlisting}
\caption{\texttt{fix\_gc} for abstract garbage collection} \label{fixgc}
\end{figure}

@@

\vspace{-2em}
\subsection{Modular Staged Analysis for Free} \label{modular}

@@

Consider there is a library that contains a list map function \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{map}:
\begin{lstlisting}
  (define (map xs f) (if (null? xs) '() (cons (f (car xs)) (map (cdr xs) f))))
\end{lstlisting}

The specialization (0CFA and without store-widening) of \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{close} on
\lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{map} generates the following next-stage Scala code, that is, the
function \lstinline[keywordstyle=,flexiblecolumns=false,mathescape=false,basicstyle=\tt]{x39}:
\begin{lstlisting}
  val x39 = {(x40: List[Set[AbsValue]],          // argument value list
              x41: Map[Addr, Set[AbsValue]],     // store
              x42: Map[Config, Set[ValSt]],      // in cache
              x43: Map[Config, Set[ValSt]]) =>   // out cache
    val x50 = List[Addr](ZCFAAddr("xs"), ZCFAAddr("f")) // addresses of arguments
    val x51 = x50.zip(x40)                              // addresses paired with their values
    val x61 = x51.foldLeft (x41) { case (x52, x53) =>   // prepare a new store that joins
      val x54 = x53._1;  val x55 = x53._2               //   the arguments and their values
        val x58 = x52.getOrElse(x54, x57)               //   into the latest store, initially x41
        val x59 = x58.union(x55);  x52 + (x54 -> x59) } 
    ... }
\end{lstlisting}

We borrow the notation traditional modular analysis
\cite{DBLP:conf/cc/CousotC02} and realize that such generated function is a
reusable artifact which can be used modularly.  The generated functions already
eliminate the interpretation overhead and monadic overhead, therefore can be
consider as partial summaries. Although, we still need to run the generated
function at next stage to obtain the analysis result.  At next stage, we can
provide different abstract arguments, store and caches, depending on the
context where it is called, thus the analysis is modular.

### 2019-08-16 paper/70_eval.tex
update and spaces

\paragraph{Implementation and Evaluation Environment}
We implement the abstract interpreters for control-flow analysis, i.e., a
value-flow analysis that computes which lambda terms can be called at each
call-sites in functional languages. With other suitable abstract domains, the
abstract interpreter also can be extended to other analyses. A front-end
desugars Scheme to a small core language that makes the analyzer easy to
implement. As described in Section \ref{cfa}, we implement two monovariant
0-CFA-like analysis, one is equipped with store-widening, the other does not.
\todo{why store widening?} Our prototype implementation currently generates
Scala code; the generated Scala code will be compiled and executed also on JVM.
it is possible to also generate C/C++ code in the future. In the experiments, we
implement several optimizations mentioned in Section \ref{staged_ds},
specifically, the selective caching and rewriting rules exploiting
partially-static data. These optimizations are useful to reduce the size of
generated code.

@@

\begin{figure}[h]
\footnotesize
\begin{tabular}{@{}ll|lll|lll@{}}
\toprule
    program             &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ & unstaged   & staged    & $\frac{\text{unstaged}}{\text{staged}}$  \\ 
    \midrule
                        &      & \multicolumn{3}{c}{w/o store-widening}  &  \multicolumn{3}{c}{w/ store-widening}\\
    \midrule
    @fib@               & 32   & 3.288 ms   & 0.154 ms   & 21.33x      & 1.434 ms   & 0.098 ms  &  14.62x       \\
    @rsa@               & 451  & 238.171 ms & 23.333 ms  & 10.20x      & 11.977 ms  & 1.197 ms  &  10.00x       \\
    @church@            & 120  & 61.001 s   & 4.277 s    & 14.26x      & 2.338 ms   & 0.534 ms  &  4.37x        \\
    @fermat@            & 310  & 23.540 ms  & 2.885 ms   & 8.05x       & 7.146 ms   & 0.915 ms  &  7.81x        \\
    @mbrotZ@            & 331  & 665.456 ms & 66.070 ms  & 10.07x      & 11.008 ms  & 1.476 ms  &  7.45x        \\
    @lattice@           & 609  & 29.230 s   & 2.627 s    & 11.12x      & 16.432 ms  & 2.427 ms  &  6.76x        \\
    @kcfa-worst-16@     & 182  & 44.431 ms  & 3.211 ms   & 13.83x      & 4.425 ms   & 0.850 ms  &  5.20x        \\
    @kcfa-worst-32@     & 358  & 284.268 ms & 9.065 ms   & 31.35x      & 10.109 ms  & 1.661 ms  &  6.08x        \\
    @kcfa-worst-64@     & 710  & 2.057 s    & 0.029 s    & 70.23x      & 23.269 ms  & 3.312 ms  &  7.02x        \\
    @solovay-strassen@  & 523  & 5.078 s    & 0.766 s    & 6.62x       & 18.757 ms  & 3.142 ms  &  5.96x       \\
    @regex@             & 550  & -          & -          & -           & 6.803 ms   & 1.088 ms  &  6.24x       \\
    @matrix@            & 1732 & -          & -          & -           & 85.611 ms  & 9.297 ms  &  9.20x       \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for monovariant control-flow analysis.} \label{evaluation_result}
\end{figure}

Figure~\ref{evaluation_result} shows the evaluation result, comparing
the performance improved by staging on two monovariant CFA (without
and with store-widening). The \textit{unstaged} and \textit{staged}
columns show the median time to finish the analysis. The column
$\frac{\text{unstaged}}{\text{staged}}$ shows the improvement from
staging. The dash - in some cells indicates timeout.  As we can see
from the table, the staged version significantly outperforms the
unstaged version on all benchmarks.  On average, the staging produces
code that runs 17x times faster than the unstaged analyzer.

\todo{analyze some interesting examples}

Our evaluation result provides an abstraction without regret
approach to construct optimizing abstract interpreters -- the user may write a
high-level modular and compositional abstract interpreter, and then derive a
staged version in a principled way that runs significantly faster.

>>>>

\paragraph{Implementation and Evaluation Environment}
We implement the abstract interpreters for control-flow analysis, i.e., a
value-flow analysis that computes which lambda terms can be called at each
call-sites in functional languages. With other suitable abstract domains, the
abstract interpreter also can be extended to other analyses. A front-end
desugars Scheme to a small core language that makes the analyzer easy to
implement. We implement two common monovariant 0-CFA-like analysis, one is
equipped with store-widening (0CFA-SW for short), the other does not (0CFA for
short). The store-widened analysis is expected to be faster than the one
without.  Our prototype implementation currently generates Scala code; the
generated Scala code will be compiled and executed on JVM. it is possible to
generate C/C++ code in the future. In the experiments, we implement several
optimizations mentioned in Section \ref{staged_ds}, specifically, the selective
caching scheme and IR-level rewriting rules (for pairs, lists, sets, and maps) exploiting
partially-static data. These optimizations are useful to reduce the size of
generated code.

@@

\vspace{-1em}
\begin{figure}[h]
\footnotesize
\begin{tabular}{@{}ll|lll|lll@{}}
\toprule
    program             &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ & unstaged   & staged    & $\frac{\text{unstaged}}{\text{staged}}$  \\ 
    \midrule
                        &      & \multicolumn{3}{c}{w/o store-widening}  &  \multicolumn{3}{c}{w/ store-widening}\\
    \midrule
    @fib@               & 32   & 3.288 ms   & 0.154 ms   & 21.33x      & 1.434 ms   & 0.098 ms  &  14.62x       \\
    @rsa@               & 451  & 238.171 ms & 23.333 ms  & 10.20x      & 11.977 ms  & 1.197 ms  &  10.00x       \\
    @church@            & 120  & 61.001 s   & 4.277 s    & 14.26x      & 2.338 ms   & 0.534 ms  &  4.37x        \\
    @fermat@            & 310  & 23.540 ms  & 2.885 ms   & 8.05x       & 7.146 ms   & 0.915 ms  &  7.81x        \\
    @mbrotZ@            & 331  & 665.456 ms & 66.070 ms  & 10.07x      & 11.008 ms  & 1.476 ms  &  7.45x        \\
    @lattice@           & 609  & 29.230 s   & 2.627 s    & 11.12x      & 16.432 ms  & 2.427 ms  &  6.76x        \\
    @kcfa-worst-16@     & 182  & 44.431 ms  & 3.211 ms   & 13.83x      & 4.425 ms   & 0.850 ms  &  5.20x        \\
    @kcfa-worst-32@     & 358  & 284.268 ms & 9.065 ms   & 31.35x      & 10.109 ms  & 1.661 ms  &  6.08x        \\
    @kcfa-worst-64@     & 710  & 2.165 s    & 0.0425 s   & 50.85x      & 23.269 ms  & 3.312 ms  &  7.02x        \\
    @solovay-strassen@  & 523  & 5.078 s    & 0.766 s    & 6.62x       & 18.757 ms  & 3.142 ms  &  5.96x       \\
    @regex@             & 550  & -          & -          & -           & 6.803 ms   & 1.088 ms  &  6.24x       \\
    @matrix@            & 1732 & -          & -          & -           & 85.611 ms  & 9.297 ms  &  9.20x       \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for monovariant control-flow analysis.} \label{evaluation_result}
\end{figure}

Figure~\ref{evaluation_result} shows the evaluation result, comparing
the performance improved by staging on the two monovariant CFAs 
The \textit{unstaged} and \textit{staged} columns show the median time to
finish the analysis. The column $\frac{\text{unstaged}}{\text{staged}}$ shows
the improvement from staging. The dash - in some cells indicates timeout.  As
we can see from the table, the staged version significantly outperforms the
unstaged version on all benchmarks.
For 0CFA w/o store-widening, we observe an average speed-up of ~17x times.  For
0CFA w/ store-widening, the average speed-up is ~7x times.  The @kcfa-worst-64@
benchmark has the most speed-up among all the tests. Although it has a
comparablely large size of AST, it has a quite simple structure which only
involves booleans, functions and applications. We conjecture that the simple
domain and structure allow the generated code to be efficiently optimized by
the JVM. But on the corresponding 0CFA-SW, we do not observe such great
speed-up. Due to the complexity of generated code and next-stage running
platform (JVM), we leave the detailed empirical analysis of which static
analysis can benefit more from specialization as a future work.

%Our evaluation result provides an abstraction without regret
%approach to construct optimizing abstract interpreters -- the user may write a
%high-level modular and compositional abstract interpreter, and then derive a
%staged version in a principled way that runs significantly faster.
